HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-07-16 scnn.tex
invariance

Cosmological and astrophysical data often comes in the form of spherical sky maps.
Observables that cover large parts of the sky, such as the Cosmic Microwave Background (CMB) \citep{planck2015cosmologicalparameters,komatsu2011sevenyear,staggs2018recentdiscoveries}, neutral hydrogen \citep{santos2015cosmologySKA,HI4PI2016fullskyHI}, galaxy clustering \citep{alam2017clusteringgalaxies}, gravitational lensing \citep{troxel2017darkenergy,hildebrandt2017kidscosmological}, and others, have been used to constrain cosmological and astrophysical models. Deep Learning has recently gained popularity as an analysis tool in cosmology \citep{schmelze2017cosmologicalmodel,luciesmith2018machinelearning,gupta2018nongaussianinformation,gillet2018deeplearning,hassan2018reionizationmodels,aragoncalvo2018classyfyinglarge,ciuca2017cnnstring}, thanks to its ability to capture complicated, non-linear patterns which are often present in the data. Convolutional Neural Networks (CNNs) are particularly well suited for analysis of astrophysical data due to the translation equivariance built into their architecture. Indeed, sky maps are translation equivariant, which means that rotating maps on the sphere doesn't change their interpretation (at least at the micro scale). As such, learned weights can be shared across the sphere: the network does not have to relearn to detect objects or features at every spatial location.

@@

% cohen2017convolutional (ICML workshop) and cohen2018spherical (ICLR best paper) are redundant
A first attempt to generalize CNNs to spherical data was to apply a standard 2D CNN to a grid discretization of the sphere~\citep{boomsma2017spherical}. While this approach leverages the well-developed 2D convolution and hierarchical pooling, it is restricted to those very specific grid samplings.
A second attempt by~\citet{cohen2018spherical} was to leverage the spherical Fourier transform and to perform the convolution in the spectral domain, thanks to the convolution theorem. While there exists fast spherical Fourier transforms for some samplings, this approach is still computationally expensive (see \figref{filtering_speed}).
Finally, the use of a graph to model the discretized sphere was considered by~\citet{khasanova2017graph} as well, towards the goal of building invariances in the network. As~\citep{boomsma2017spherical}, they use an equirectangular grid sampling. Moreover, they did not take advantage of a hierarchical pixelization of the sphere and resorted to dynamic pooling~\citep{kalchbrenner2014dcnn}.

@@

The use of a graph to model the discretized sphere was considered by~\citet{khasanova2017graph} as well. The authors map omnidirectional images to the sphere, with the goal of correcting for distortions induced by the lens used for the acquisition.
Besides the different application covered by the paper, there are two important
differences. First the sampling is different as~\cite{khasanova2017graph} uses a
grid. Second, it uses a different graph CNN based on a different parameterization
of the convolution kernel.

>>>>

Cosmological and astrophysical data often comes in the form of spherical sky maps.
Observables that cover large parts of the sky, such as the Cosmic Microwave Background (CMB) \citep{planck2015cosmologicalparameters,komatsu2011sevenyear,staggs2018recentdiscoveries}, neutral hydrogen \citep{santos2015cosmologySKA,HI4PI2016fullskyHI}, galaxy clustering \citep{alam2017clusteringgalaxies}, gravitational lensing \citep{troxel2017darkenergy,hildebrandt2017kidscosmological}, and others, have been used to constrain cosmological and astrophysical models. Deep Learning has recently gained popularity as an analysis tool in cosmology \citep{schmelze2017cosmologicalmodel,luciesmith2018machinelearning,gupta2018nongaussianinformation,gillet2018deeplearning,hassan2018reionizationmodels,aragoncalvo2018classyfyinglarge,ciuca2017cnnstring}, thanks to its ability to capture complicated, non-linear patterns which are often present in the data. Convolutional Neural Networks (CNNs) are particularly well suited for analysis of cosmological data due to the translation equivariance built into their architecture. Indeed, sky maps are translation equivariant, which means that rotating maps on the sphere doesn't change their interpretation. Only the statistics of the maps is relevant. The location of the map on the sky is always fixed (by where the telescope was looking), and most of the models tested are homogeneous and isotropic (the cosmological principle). As such, learned weights can be shared across the sphere: the network does not have to relearn to detect objects or features at every spatial location.

@@

% cohen2017convolutional (ICML workshop) and cohen2018spherical (ICLR best paper) are redundant
A first attempt to generalize CNNs to spherical data was to apply a standard 2D CNN to a grid discretization of the sphere~\citep{boomsma2017spherical}. While this approach leverages the well-developed 2D convolution and hierarchical pooling, it is restricted to those very specific grid samplings which are highly irregular on the sphere.
A second attempt by~\citet{cohen2018spherical} was to leverage the spherical Fourier transform and to perform the convolution in the spectral domain, thanks to the convolution theorem. While there exists fast spherical Fourier transforms for some samplings, this approach is still computationally expensive (see \figref{filtering_speed}).
Finally, the use of a graph to model the discretized sphere was considered by~\citet{khasanova2017graph} as well, towards the goal of building invariances in the network. While their architecture is fully invariant to translation and rotation on the sphere, they did not take advantage of a hierarchical pixelization of the sphere, and the statistical layer used to be invariant needs to be hand-crafted to capture statistics of interest.

@@

The use of a graph to model the discretized sphere was considered by~\citet{khasanova2017graph} as well. The authors map omnidirectional images to the sphere, with the goal of correcting for distortions induced by the lens used for the acquisition.
Besides the different application covered by the paper, there are two important
differences. First the sampling is different as~\cite{khasanova2017graph} uses a
grid. Second, it uses a different graph CNN based on a different parameterization
of the convolution kernel.
As~\citep{boomsma2017spherical}, they use an equirectangular grid sampling. Moreover, they did not take advantage of a hierarchical pixelization of the sphere and resorted to dynamic pooling~\citep{kalchbrenner2014dcnn}.
Finally, while we do want our network to be rotation invariant, their statistical layer --- an operation which computes a set of spatially invariant statistics of the last feature map --- destroys a lot of information \todo{and hinders performance (should be empirically verified)}. While being completely invariant is a nice goal, CNNs for image recognition (arguably the most researched and developed architectures), are not built to be invariant to translation. As with the proposed spherical CNN, the fully connected layers at the end of the network depend on the spatial localization of features. While that may not be ideal, it was shown to work well in practice.
